---
title: "lab-7-working"
author: "Vienna Saccomanno"
date: "2/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tmap)
library(sf)
library(spatstat)
library(maptools)
library(sp)
library(raster)
# library(rgdal)
library(gstat)
library(plotKML) # for points to raster (they won't have this...just testing)
```

###Part 1. Hawaii raster intro

```{r}

# Read in the raster data because sf is only set up for vector data

hi_par <- raster("PAR_CLIM_M.tif")
hi_sst <- raster("SST_LTM.tif")
hi_chl <- raster("CHL_LTM.tif")
  
# Base plots
plot(hi_par)
plot(hi_sst)
plot(hi_chl)

par(mrow = c(1,3))
plot(hi_par)
plot(hi_sst)
plot(hi_chl)

#hi_sst - tells you about the data = raster

```

First: some useful functions for rasters

Checking it out: 

- crs
- reprojection
- cropping
- simple algebra example


```{r}

hi_sst@crs # Shows CRS(coordinate reference system): NAD83

hi_sst@extent # Shows extent (bounds)

```

Example: reprojection to WGS84
```{r}
wgs84 = "+proj=longlat +datum=WGS84 +ellps=WGS84 +no_defs" # Projection information, Just have this ready to copy/paste

# Reproject hi_sst to be WGS84. Biliner = liner interpertation in to directions
hi_sst_84 = projectRaster(hi_sst, crs = wgs84, method = "bilinear")

# Check the reprojection - looks good
hi_sst_84@crs
hi_sst_84@extent #now the lat long are in familiar units

```

raster::aggregate() for resampling to make files size smaller
```{r}

# Sea surface temperature: 
sst_rs <- aggregate(hi_sst, fact = 10) #Decreasing resolution by order of 10, Quick way to resample. Raster data files can get super massive and if you just want to test code, might want to resample to make things quicker.

plot(sst_rs)

# Plot side-by-side for comparison:
par(mfrow = c(1,2))
plot(hi_sst)
plot(sst_rs)

```

Crop a raster: 
```{r}

#what are the current extents?, get the bounding box for the raster
hi_sst_84@extent

# Get these extents from hi_sst_84 (call in console to see) what the actual limits are for hi_sst_84, then decide on cropping boundaries. Allison chose a Hawaiian island for this lab.

# First create a spatial polygon. as() forces this to be a spatial polygon.
bounds <- as(extent(-156.2, -154.5, 18.7, 20.4), 'SpatialPolygons') # Keep in mind, this could be any polygon shape (state outline, county outline, etc.) 'spatialPolygons', tells R to convert to spatial polygons. We're cropping to a rectangle, but we could have pulled in shapefile data and cropped accordingly.

# Reproject - if no projection this is the easy way to do this. Making crs of "bounds" match the crs of hi_sst_84. Need them to match in order to do math with them. Bounds = polygon, hi_sst_84= raster.
crs(bounds) <- crs(hi_sst_84)

# Then crop the raster by the polygon bounds 
sst_crop <- crop(hi_sst_84, bounds)

# And plot:
plot(sst_crop)
```


A simple algebra example/ raster math: 

Let's say we're creating a nonsensical variable called "tropicality", which is the sum of the PAR + SST + 2*ChlA. How can we create a layer for tropicality? How do we map?

First let's reprojeect PAR (radiation) and ChlA and get everything into the same CRS:

Use method = "bilinear" for continuous variables, "ngm" /nearest neighbor for categorical
```{r}

#wgs84 is stored from above and we can reuse it here. Bilinear is the default for continuous
hi_par_84 <- projectRaster(hi_par, crs = wgs84, method = "bilinear")

hi_chla_84 <- projectRaster(hi_chl, crs = wgs84, method = "bilinear")

# Now we have PAR, Chl-a, and SST all in the same CRS (WGS84) and can start doing some simple algebra. 
```
Plot them side-by-side:
```{r}
par(mfrow = c(1,3))
plot(hi_sst_84)
plot(hi_chla_84)
plot(hi_par_84)

#Checking out the scales of the three variables. This way we know the potentail ranges of adding our raster data together (~75)
```


Raster math is pretty straightforward: 
```{r}
#It says raster extents are different - not perfectly additive. Only intersections included. Sometimes will need to coerce raster data to be perfectly aligned.

#Raster math challenges: different extents/resultions/projections, and big datasets

trop <- hi_par_84 + hi_sst_84 + 2*hi_chla_84
plot(trop)
```


And we might want to plot these in tmap instead: 

Let's look at sea surface temperature. 
```{r}

islands <- read_sf(dsn = 'islands', layer = "Island_boundaries") %>% #file name + file type
  dplyr::select(Island) %>% #only select island name. Make sure to call dplyr to use select form dplyr
  st_simplify(dTolerance = 10) %>% 
  st_transform(crs = 4326) #need to transform to make sure same crs

plot(islands)

tmap_mode("plot") # or switch to tmap_mode("view")

tm_shape(hi_sst_84) + 
  tm_raster(title = "Mean Sea Surface Temperature") + #raster layer first
  tm_layout(bg.color = "navyblue", #change background color
            legend.position = c("left","bottom"),
            legend.text.color = "white", 
            legend.text.size = 0.5) +
  tm_shape(islands) +
  tm_fill("darkgreen")

# Or name it and export
sst_map <- tm_shape(hi_sst_84) +
  tm_raster(title = "Mean Sea Surface Temperature") +
  tm_layout(bg.color = "navyblue", 
            legend.position = c("left","bottom"),
            legend.text.color = "white", 
            legend.text.size = 0.5) +
  tm_shape(islands) +
  tm_fill("darkgreen")

#Saving - will show up in the working directory
tmap_save(sst_map, "sst.png", height=5)

```

What if only want to retain parts of a raster?
Example: Conditional rasters and masking

Let's say we have a sensitive species and we're trying to find suitable habitat/ a place they like. They like warm water (average temp >= 25.6 deg C) and PAR below 54.
```{r}
# Call in console. Currently don't have matching extents, columns nor rows, we need to update:
extent(hi_sst_84) <- extent(hi_par_84)
compareRaster(hi_sst_84, hi_par_84) #just says different number of columns, becuase we just set extents
# Check compareRaster...nope. Mismatching columns & rows is still a problem. 

# But we also need to make sure they have the same number of rows & columns. So make a raster from scratch to match hi_par_84 (pull info form the console)
cr <- raster(nrow = 822, 
             ncol = 1229, 
             xmn = -160.4365, 
             xmx = -154.5373, 
             ymn = 18.7309, 
             ymx = 22.44634)

#Now we have a raster that has the same number of columns and rows as par, now can resample hi_sst_84 to match cr. Continuous variable = bilinear interpolation to resample and fill in the old columns
sst_new <- resample(hi_sst_84, cr, method = "bilinear")

compareRaster(sst_new, hi_par_84) # TRUE! = they match
```

Create cropped versions of just Kauai
```{r}
# Created 'bounds_main' as earlier: 

bounds_main <- as(extent(-159.9, -159.2, 21.7, 22.3), 'SpatialPolygons') # Keep in mind, this could be any polygon shape (state outline, county outline, etc.)

# Reproject
crs(bounds_main) <- crs(sst_new)

par_kauai <- crop(hi_par_84, bounds_main) #croping the PAR raster to fit Kauii
sst_kauai <- crop(sst_new, bounds_main) #cropping SST raster to fit Kauii

# Check out PAR:
plot(par_kauai)

# Then SST:
plot(sst_kauai)

```


Subsetting and setting everything else to NA
Now we only want to isolate regions where the temperature >= 25.4 and PAR < 54.
```{r}
# Habitat
par_hab <- par_kauai # just makes a copy
par_hab[par_hab >= 54.0] <- NA #take PAR raster info and subset every PAR value above 54 to NA

plot(par_hab) #only hab less than 54 shows up!

sst_hab <- sst_kauai # also makes a copy
sst_hab[sst_hab < 25.6] <- NA #temps greater than or equal to 35.6

plot(sst_hab)

#where do these overlap? Using raster::mask
suit_hab <- mask(sst_hab, par_hab)
plot(suit_hab)

par(mfrow = c(1,2))
plot(par_hab)
plot(sst_hab)

```
